---
title: "Simulated Data: Nurse Staffing and Quality of Care"
author: "Nikhil Haas"
date: "May 13, 2015"
output: pdf_document
---

# Introduction

This simulated data set is based on the journal article "Is More Better? The
Relationship between Nurse Staffing and the Quality of Nursing Care in
Hospitals" in Medical Care (Sochalski, 2004). In this study, Sochalski
analyzed data obtained from a 1999 survey of 8,670 inpatient nurses from
Pennsylvania, U.S.A. hospitals. The survey asked nurses to rate the
quality of care from their last shift and estimate the number
of patients they cared for, the number of tasks that went undone, and the
number of patient safety problems that occurred during the same shift. The
author found several correlations between the responses to the questions. The
questions and response data from survey are the inspiration for this simulated
data set.

# Method

Here, we generate artificial responses to the questions that were in the 1999
survey. In this simulated scenario, the responses are entered into a hospital
administrative system over several years by its nurses in an effort to better
understand what the hospital can do to improve patient care. The hospital has
no knowledge of what factors correlate with patient care and is hoping to use
the data obtained to discover relationships that will help them achieve their
goal of delivering better care.

## Questions and Response Ranges

Nurses have been asked to complete the survey following each shift. Below are
the columns of data that will hold the response data and the expected ranges
for each response type.

1. Employee ID: Between 0 and 9000
2. Quality of care as rated by nurses: Poor, Fair, Good, Excellent
3. Number of registered nurses on-staff during last shift: 1 - 30
4. Number of patients cared for during last shift: 0 - 20
5. Number of tasks that went undone during last shift: 0 - 7
6. Number of patient safety problems during last shift: 0 - 4
7. Date this survey was taken: YYYY/MM/DD

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

# References

Sochalski, J. (2004). Is more better?: the relationship between nurse staffing
and the quality of nursing care in hospitals. Medical care, 42(2), II-67.
